Individual-based numerical experiment to describe the distribution of floating kelp within the Southern Benguela Upwelling System.

### 2021-10-31 thesis_chapter_4.Rmd
updated chapter 4 files

---
title: "Chapter 4"
author: "Ross Coppin"
date: "11/12/2020"
output:
  html_document: default
  word_document: default
  pdf_document: default
---

```{r message=FALSE, warning=FALSE, include=FALSE}

library(tidyverse)
library(flextable)
library(tidync)
library(ncdf4)
library(lubridate)
library(gridExtra)
library(ggpubr)

@@

The raft-time is the amount of time macroalgae floats at the surface before becoming negatively buoyant and begins to sink to the ocean-floor. Raft-time, also referred to as raft longevity, is determined by epibiont load, macroalgal growth rate and the loss of kelp biomass over time through fragmentation. Epibiont load consists of epiphytic and bryozoan species which tend to grow more rapidly in higher light and temperature environments, such as the ocean surface [@thiel2005; @thiel2005; @graiff2016]. The increase in epibiont load reduces the buoyancy of floating macroalgal rafts over time eventually causing them to sink to the seafloor, while also potentially increasing drag [@hobday2000; @tala2013; @craw2018]. Fragmentation of kelp biomass over time may also reduce raft times further, while growth may counteract the effects of epibiont load and fragmentation provided the environmental conditions are favorable [@graiff2016; @macaya2016]. For instance a study by @rothausler2011 investigated the abiotic and biotic factors influencing raft time and dispersal potential of the giant kelp *M. pyrifera* by a combination of tethering experiments and field surveys. The results showed the physiological performance of kelp declined with increasing epiphyte biomass and that higher temperatures increased growth rate of epibionts and fragmentation of kelp. The authors concluded from the results of both the experiments and field surveys that *M. pyrifera* dispersal is dependent on low temperature and moderate irradiance conditions, with high temperatures and higher irradiance reducing overall raft-time and hence dispersal potential [@rothausler2011].

@@

## Raft-time
Raft-time or sometimes referred to as raft-longevity is the time a macroalgae raft remains afloat at the surface of the ocean after detachment before sinking. Raft-time was estimated from the results of past studies by @graiff2016 and @graiff2013 for each season. For summer the raft time was 22 days, autumn was 41 days, winter was 38 days, and spring was 33 days.

```{r morphology surface area table, message=FALSE, warning=FALSE, include=TRUE, echo=FALSE, fig.cap="TBD."}

@@

# Add in and/calculate estimates
morph_data$stipe_radius <- with(morph_data, stipe_circ/(pi*2))
morph_data$frond_width <- with(morph_data, 5)
morph_data$frond_thickness <- with(morph_data, 0.2)
morph_data$bulb_length <- with(morph_data, 10)
morph_data$bulb_base_radius <- with(morph_data, 5)
morph_data$top_radius <- with(morph_data, 3)
morph_data$bottom_radius <- with(morph_data, 10)
morph_data$stipe_radius <- with(morph_data, stipe_circ/(pi*2))
morph_data$height <- with(morph_data, 5)

@@

# Make table based on new entries
site_surface_area <- morph_data
site_surface_area <- flextable(data = site_surface_area)
#site_surface_area <- theme_zebra(site_surface_area, odd_header = "transparent", even_header = "transparent")
site_surface_area <- align(x= site_surface_area, align = "center", part = "all")
#area_table <- set_header_labels(area_table, Plant.characteristic = "Plant characteristic", Approximate.shape = "Approximate shape", Equation.used = "Equation Used", Plant.dimensions = "Plant dimensions")
#area_table <- set_header_labels(area_table, Ac.Minimum = "Ac Minimum", Ac.Mean = "Ac Mean", Ac.Maximum = "Ac Maximum")
site_surface_area <- set_table_properties(site_surface_area, width = 1, layout = "autofit")
#site_surface_area <- set_caption(site_surface_area, "Table 2:Summary of the atrributes and estimates used to calculate the overall surface area of Kommetjie kelp individuals used in the various simulation.")
site_surface_area
```

@@

Simulations were run using the OceanParcels software (available from https://oceanparcels.org/). All simulations were run using a Runge-Kutte fourth order advection scheme, while custom kernels simulated the effects of hydrodynamic and wind drag on overall trajectory, seasonal raft-times, sinking and settling on seafloor of virtual kelp particles; all important processes in the determination of surface and bottom sinks. The simulations were run for 2019, which was divided into various seasons (see table??). For incorporating drag, the 'maximum exposure scenario' proposed by @coppin2021 was used for all simulations (85% hydrodynamic and 15% wind drag) while virtual kelp particles were on the surface and then would convert to full hydrodynamic drag exposure (100%) once the virtual kelp started sinking; as wind would no longer act as a direct influencer of trajectory once below the ocean surface.

@@

# Results
## Ocean analysis

```{r eval=FALSE, message=FALSE, echo=FALSE, include=FALSE}

>>>>

---
title: "Individual-based numerical experiment to describe the distribution of floating kelp within the Southern Benguela Upwelling System."
author: 
- R Coppin^[Department of Biodiversity and Conservation Biology, University of the Western Cape, Private Bag X17, Bellville 7535, South Africa], C Rautenbach^[National Institute of Water and Atmospheric Research, Hamilton: Gate 10 Silverdale Road Hillcrest, 3216], AJ Smit^[South African Environmental Observation Network, Elwandle Coastal Node, Port Elizabeth, South Africa]
output:
  word_document: 
    reference_docx: markdown/wiley_template.docx
  html_document: default
  pdf_document: default
bibliography: references.bib
csl: markdown/limnologyandoceanography.csl
---

```{r message=FALSE, warning=FALSE, include=FALSE}
library(tidyverse)
library(flextable)
library(tidync)
library(ncdf4)
library(lubridate)
library(gridExtra)
library(ggpubr)
library(jpeg)
library(gridExtra)
library(grid)
```

@@

The inter- to sub-tidal habitats of temperate shorelines are usually dominated by species of macroalgae, specifically brown seaweeds of the order Laminariales; commonly referred to as kelp [@Steneck2002;@Schiel2015-vr]. Kelps are highly productive organisms [@Mann1973;@Miller2011-pm; @Dayton1998-xw;@Krumhansl2012;@Pedersen2012-vr] which provide a 3-dimensional habitat for various fauna and flora [@anderson1997;@Leliaert2000a;@Steneck2002; @Norderhaug2012-ks; @Krumhansl2016]. Kelp are also capable of altering the local wave and light environment by wave dampening [@Mork1996-md; @Mia2013] and shading of the understorey habitat [@duggins1990; @watanabe1992; @clark2004], as well as providing important organic subsidy through erosion of fronds and fragmentation of plant structures [@Krumhansl2012; @Filbee-Dexter2018-wp; @DeBettignies2013; @Simonson2015; @ramirez2021]. Due to kelps’ ecological role in the nearshore, they are regarded as ecological engineers as well as important influencer of biodiversity along temperate shorelines [@FowlerWalker2005;@graham2004;@Harley2012;@Smale2017; @Wing2007-tv;@Wernberg2008]. Kelps are also important from a socio-economic perspective, as they provide habitats for economically important species such as abalone and rock lobster [@vasquez2014; @blamey2018], and are harvested for uses ranging from feed for the aquaculture sector [@troell2006] to pharmaceuticals [@smit2004].

Kelps’ ecological and socio-economic contributions are not limited to their local environment. During times of high wave energy kelps can become dislodged from the substrate  [@seymour1989; @Thomsen2004] and are transported via ocean currents, eventually deposited on beaches or offshore areas [@hobday2000a]. The kelp deposited on beaches provides important organic subsidies for coastal invertebrate communities, as well as a resource which can be harvested for economic purposes [@kirkman1997;@troell2006; @holden2018]. Along with the organic input are 'hitchhikers', which are organisms that remain attached during offshore transport once dislodged [@highsmith1985; @holmquist1994; @macaya2016; @thiel2005]. The transport of other organisms attached to kelp is regarded as an important processes in invasive species intrusion [@lewis2005; @McCormick2008; @nikula2010; @batista2018] and marine connectivity [@norton1992, @ingolfsson1995; @hobday2000a; @fraser2011]. The formation of kelp-rafts can also occur, providing habitat complexity in the pelagic environment for various fish species [@kingsford1992; @kingsford1995], Kelps which are transported offshore can become entangled to form a kelp-raft, but there is no clear mechanism for kelp-raft formation. The structural integrity of individual kelp or rafts will degrade over time through a combination of biotic and abiotic factors which ultimately results in sinking. Once settled on the seafloor, the kelp will degrade further over time resulting in organic input into offshore areas; which may be transported back towards the coast through upwelling processes [@Krumhansl2012; @Filbee-Dexter2018-wp; @ramirez2021].

Understanding the distribution patterns of macroalgae in the ocean and the underlying oceanographic processes influencing trajectory has received much attention in the literature. Approaches such as radio-tagged kelp rafts [@harrold1989], geological forms in holdfasts [@waters2018], genetic approaches [@craw2018; @waters2018; @nikula2013], the presence of region-specific epibionts on beach-cast kelp [@fraser2011], and Lagrangian approaches [@putman2018; @brooks2019; @putman2020] have helped to shed light on the processes involved in macroalgae transport.  Ocean currents are the main influencer of trajectory as well as drag and windage, which is the direct effect of wind on a floating object [@onink2019; @shen2019; @dasaro2018; @gates2018; @beron2016; @trinanes2016; @furnans2008]. More recent work by @coppin2021 has highlighted the importance of both hydrodynamic and wind drag in entrainment patterns within oceanographic mesoscale processes, more specifically eddies.

Although the past research concerning the transport of kelp has greatly improved our understanding of such processes, limitations and gaps still exist. For example, kelps are highly morphologically diverse which may influence buoyancy and overall drag characteristics. Morphological characteristics, sea surface temperatures, and light also influence the time which kelps remain afloat at the surface; known as raft-times. Macroalgae rafts also vary greatly in size, therefore studies investigating a specific raft size limits our understanding of that particular scenario as the findings cannot be inferred for other sizes of kelp rafts. Finally, it has been well established that morphological characteristics of macroalgae are influenced by local abiotic factors, which can vary greatly along coastlines. Although factors such as light, turbidity, and nutrients have been shown to influence macroalgae morphology; temperature and waves are regarded as the main influencers [@Hurd2000;@Denny2002a;@Thomsen2004;@FowlerWalker2005;@Wernberg2005c;@mabin2013;@bekkby2014; @coppin2021]. How the variable response in morphological characteristics influences sink rates, buoyancy, and drag is also yet to be established. There is also no current work where rafts may form on the surface and the role of eddies in this regard. It is most likely that eddies might play a role in this regard as they have shown to cause collection of other flotsam at the surface via entrainment @onink2019; @dasaro2018; @brach2018. For example, a study by @beron2019 showed that the inclusion of inertia in numerical experiments resulted in entrainment patterns of flotsam in eddies which more closely reflected observational data.

A better understanding of the distribution of floating kelp and ocean processes involved will aid in understanding the broader ecological role that kelp plays in the marine environment but will also be useful for coastal management and harvesting activities. The current work aims to shed light on the distribution patterns of floating kelp by a series of Lagrangian individual-particle tracking numerical experiments. This approach will allow for the morphological characteristics that vary with the local temperature and wave environment to be accounted for. In addition, since individual kelp has a higher probability of dislodgement compared to kelps in which holdfasts are clustered together, an individual-based approach will reflect real-world scenarios [@Thomsen2004]. The novel approach of including hydrodynamic and wind drag based on site-specific morphological characteristics by @coppin2021 will be used and the current work will build on this approach by the inclusion of morphological specific buoyancy and sinking rates.

@@

## Morphological Characteristics

Morphological characteristics of kelps are influenced by local wave and temperature conditions, which in turn will influence buoyancy, drag, and sinking parameters. Kelp morphological data was garnered from a previous study by @coppin2020. The assumption made is that the largest kelps (largest surface area) are the most likely to be dislodged from the substrate, therefore only maximum surface areas were garnered from the data.  In instances where the morphological data needed were not available, reasonable estimates were used based on previous studies and experience. To calculate the surface area and volume for kelps from each site, geometric shapes were used which most closely matched the shape of the relevant kelp morphological characteristics as per @coppin2021. The shapes assigned to each of the morphological features were used in the calculation of buoyancy, drag, and sinking rate.

## Buoyancy

To effectively model the trajectory of a floating object, buoyancy needs to be calculated. The buoyancy will determine the amount of exposure to, and ultimately magnitude, of hydrodynamic and wind drag which in turn influences the trajectory of the object to the variable current and wind fields. The portion of an object above and below the surface of the water is determined by the net effect of forces acting on the object, namely buoyancy force (BF) and the weight of the object. The BF is the force pushing the object out of the water and the weight is the force pulling the object down. In order to calculate the buoyancy force (BF) the density and volume of the object must be known, such that: $$BF = p_fV_sg$$

where $p_f$ is the density of the liquid, $V_s$ is the volume of the object submerged, and $g$ is the gravitational acceleration (m/s).

The force resulting from the weight of an object can be calculated by $W_f = p_oV_og$. Substitution of mass for density and volume alters the equation to: $$W_f = p_oV_og$$
where $p_o$ is the density of the object, $V_o$ is the total volume of the object and $g$ is gravitational acceleration. An object will only float if the forces acting on it ($BF$ and $W_f$) are equal to one another, such that:  
$$BF = W_f$$
then substituting  density and volume results in:
$$p_fV_sg = p_oV_og$$
which can also be further expressed as:
$$V_s/V_o = p_o/p_f$$

The solution of equation (3) results in a unitless fraction that can be expressed as a percentage of the relevant portion submerged. This in turn can be used to calculate the percentage of the object above the surface of the water (exposed to wind). The percentage of the object below and above the water surface was used in the calculation of hydrodynamic and wind drag when the virtual kelp is at the surface. Since morphological characteristics are influenced by the local wave and temperature environment, the buoyancy of virtual kelp particles may vary from site to site.

## Sink Rate

Kelp floating on the surface of the ocean will begin to sink once the structural integrity of air-filled structures of the plants fails and becomes flooded. It is also possible for sinking to begin due to an increase in epibiont biomass which overwhelms the buoyancy of kelp, however, this is beyond the scope of this study. The target species of this study, *E. maxima*, contains a hollow stipe and a single air-filled pneumatocyst which serve to keep the plant upright in the water column, allowing the light-saturated surface waters to be reached. These structures also cause the plant to float if dislodged and therefore was taken into account when calculating the overall volume of the plant; which in turn was used to calculate other important parameters (buoyancy, drag, and sink rate).

## Vertical Velocity

The force responsible for the kelp to begin to sink once the structural integrity has been compromised is the weight of the overall plant. The calculation of the force resulting from the weight was based on previous work by @dietrich1982, @lobelle2021 and @kooi2017, such that

$$F_w = (p_s - p_l)gV$$

Where $p_s$ is the density of the particle, $p__l$ the density of the medium, $g$ is the gravitational acceleration constant, and $V$ is the total volume of the particle. In order to accurately calculate the density of the kelp based on volume and mass, the volume of the hollow section of the stipe was subtracted from the overall volume. The target species used for this experiment (*E. maxima*) has an air-filled stipe and therefore the hollowness needs to be accounted for when calculating the overall volume of the stipe. The volume of the kelp is:
$$Volume = V_t - V_h$$
where $V_t$ is the total volume of the stipe if now hollowness is taken into account and $V_h$ is the hollowness volume of the stipe.
The other component of the vertical velocity is the drag force in the opposite direction to the downward force of the weight. The vertical drag component is calculated by the same equation for drag in the horizontal, such that:

$$F_d = 1/2pv^2C_DA_c$$
where $p$ is the density of the medium, $v$ is the velocity at a particular time-step, $C_d$ is the shape drag-coefficient, and $A_c$ is the cross-sectional area of the object. However, unlike surface drag where the velocity is interpolated from the underlying ocean model, the velocity component for the vertical drag force is calculated from the same equation used for hydrodynamic and wind drag. The approach for modeling the 'behavior' of kelp was used in this paper allows for the incorporation of site-specific morphological characteristics.

Simulations were run using the OceanParcels software (available from https://oceanparcels.org/). All simulations were run using a Runge-Kutte fourth order advection scheme, while custom kernels simulated the effects of hydrodynamic and wind drag on overall trajectory, seasonal raft-times, sinking and settling on seafloor of virtual kelp particles; all important processes in the determination of surface and bottom sinks. The simulations were run for 2019, which was divided into various seasons (see table 2). For incorporating drag, the site specific buoyancy calculations were used to determine the magnitude of hydrodynamic (portion below surface) and wind (portion above the surface) while virtual kelp particles were on the surface and then converted to full hydrodynamic drag exposure (100%) once the virtual kelp started sinking; as wind would no longer act as a direct influencer of trajectory once below the ocean surface.

## Raft-time

Raft-time or sometimes referred to as raft-longevity is the time a macroalgae raft remains afloat at the surface of the ocean after detachment before sinking. Raft-time, also referred to as raft longevity, is determined by epibiont load, macroalgal growth rate and the loss of kelp biomass over time through fragmentation. Epibiont load consists of epiphytic and bryozoan species which tend to grow more rapidly in higher light and temperature environments, such as the ocean surface [@thiel2005; @thiel2005; @graiff2016]. The increase in epibiont load reduces the buoyancy of floating macroalgal rafts over time eventually causing them to sink to the seafloor, while also potentially increasing drag [@hobday2000; @tala2013; @craw2018]. Fragmentation of kelp biomass over time may also reduce raft times further, while growth may counteract the effects of epibiont load and fragmentation provided the environmental conditions are favorable [@graiff2016; @macaya2016]. For instance a study by @rothausler2011 investigated the abiotic and biotic factors influencing raft time and dispersal potential of the giant kelp *M. pyrifera* by a combination of tethering experiments and field surveys. The results showed the physiological performance of kelp declined with increasing epiphyte biomass and that higher temperatures increased growth rate of epibionts and fragmentation of kelp. The authors concluded from the results of both the experiments and field surveys that *M. pyrifera* dispersal is dependent on low temperature and moderate irradiance conditions, with high temperatures and higher irradiance reducing overall raft-time and hence dispersal potential [@rothausler2011]. Raft-time was estimated from the results of past studies by @graiff2016 and @graiff2013 for each season. For summer the raft time was 22 days, autumn was 41 days, winter was 38 days, and spring was 33 days.

```{r fieldworkdata and table wrangle, message = FALSE, echo=FALSE, warning=FALSE, include=FALSE}
### data for stipe hollowness
hollow_df <- read.csv(file = "data/chapter_4/hollowness_data.csv", sep = ";")
hollow_df$inner_radius <- with(hollow_df, (outer_diameter/2)) # addinmg difference between measurements
hollow_df$outer_radius <- with(hollow_df, (inner_diameter/2)) # addinmg difference between measurements
hollow_df$difference <- with(hollow_df, (outer_diameter - inner_diameter)) # addinmg difference between measurements
hollow_df$percentage <- with(hollow_df, (inner_diameter/outer_diameter)) # inner as percentage of outer
hollow_df_mean <- hollow_df %>% # calculate mean trajectory
  group_by(site) %>%
  summarise(across(outer_diameter:percentage, mean, na.rm= TRUE))

ggplot(data = hollow_df, aes(x = outer_diameter, y = inner_diameter)) +
  geom_point() +
  stat_smooth(method=lm)

ggplot(data = hollow_df, aes(x = no, y = percentage)) +
  geom_point() +
  geom_smooth()

ggplot(data = hollow_df, aes(x = no, y = difference)) +
  geom_point() +
  geom_smooth()

# testing for correlation
outer_df <- hollow_df$outer_diameter
inner_df <- hollow_df$inner_diameter

outer_rad <- hollow_df$outer_radius
inner_rad <- hollow_df$inner_radius

cor(outer_df, inner_df, method = c("pearson"))
cor.test(outer_df, inner_df, method = c("pearson"))
cor(outer_rad, inner_rad, method = c("pearson"))
cor.test(outer_rad, inner_rad, method = c("pearson"))

cor(outer_df, inner_df, method = c("spearman"))
cor.test(outer_df, inner_df, method = c("spearman"))
cor(outer_rad, inner_rad, method = c("spearman"))
cor.test(outer_rad, inner_rad, method = c("spearman"))

@@

# Add in and/calculate estimates
morph_data$stipe_radius <- with(morph_data, stipe_circ/(pi*2))
morph_data$frond_width <- with(morph_data, 5)
morph_data$frond_thickness <- with(morph_data, 0.2)
morph_data$bulb_length <- with(morph_data, 10)
morph_data$bulb_base_radius <- with(morph_data, 5)
morph_data$top_radius <- with(morph_data, 3)
morph_data$bottom_radius <- with(morph_data, 10)
morph_data$stipe_radius <- with(morph_data, ((stipe_circ/pi)/2))
morph_data$hollow_radius <- with(morph_data, (0.61*stipe_radius))
morph_data$hollow_length <- with(morph_data, (0.90*stipe_length))
morph_data$height <- with(morph_data, 5)
morph_data$hollow_volume <- with(morph_data, pi*(hollow_radius*hollow_radius)*hollow_length)

@@

# Calculate actual (True) kelp volume (hollowness taken into account)

morph_data$kelp_volume <- with(morph_data, (total_volume - hollow_volume))

# Calculate the density of each kelp

morph_data$kelp_density <- with(morph_data, total_mass/kelp_volume)

# Make table based on new entries
site_surface_area <- morph_data
site_surface_area <- flextable(data = site_surface_area)
site_surface_area <- theme_zebra(site_surface_area, odd_header = "transparent", even_header = "transparent")
site_surface_area <- align(x= site_surface_area, align = "center", part = "all")
site_surface_area <- theme_vanilla(site_surface_area)
site_surface_area <- fontsize(site_surface_area, size = 8, part = "body")
site_surface_area <- fontsize(site_surface_area, size = 8, part = "header")
site_surface_area <- colformat_num(site_surface_area,digits = 1)
site_surface_area <- set_header_labels(site_surface_area, frond_mass = "Frond mass", primary_length = "Primary blade length", primary_width = "Primary blade width", frond_length = "Frond length", stipe_mass = "Stipe mass", stipe_length = "Stipe length", stipe_circ = "Stipe circumference", tufts = "No. of Tufts", epi_length = "Epiphyte length", total_length = "Total length", stipe_radius = "Stipe radius", frond_width = "Frond width", frond_thickness = "Frond thickness", bulb_length = "Bulb length", bulb_base_radius = "Bulb base radius", top_radius = "Top radius", bottom_radius = "Bottom radius", hollow_radius = "Hollow radius", hollow_length = "Hollow length", height = "Height", hollow_volume = "Hollow Volume", total_Ac = "Total Ac", total_mass = "Total mass", total_volume = "Total volume", kelp_volume = "Kelp Volume", kelp_density = "Kelp density")
site_surface_area <- set_table_properties(site_surface_area, layout = "fixed")
```

```{r morphology surface area table, message=FALSE, warning=FALSE, include=TRUE, echo=FALSE}
site_surface_area <- set_caption(site_surface_area, "Table 1:Summary of the attributes and estimates used to calculate the overall site-specific surface area of kelp individuals used in the various simulations.") 
#site_surface_area <- set_table_properties(site_surface_area, layout = "autofit")
site_surface_area

@@

Simulations were run using the OceanParcels software (available from https://oceanparcels.org/). All simulations were run using a Runge-Kutte fourth order advection scheme, while custom kernels simulated the effects of hydrodynamic and wind drag on overall trajectory, seasonal raft-times, sinking and settling on seafloor of virtual kelp particles; all important processes in the determination of surface and bottom sinks. The simulations were run for two years, which were divided into the relevant seasons seasons (see table 2).

```{r simulation table, message=FALSE, warning=FALSE, include=TRUE, echo=FALSE, fig.cap="TBD."}

chapter4_table2 <- read_csv2(file = "tables/chapter_4/simulation_table.csv")
chapter4_table2 <- flextable(data = chapter4_table2)
chapter4_table2 <- theme_zebra(chapter4_table2, odd_header = "transparent", even_header = "transparent")
chapter4_table2 <- align(x= chapter4_table2, align = "center", part = "all")
chapter4_table2 <- theme_vanilla(chapter4_table2)
chapter4_table2 <- fontsize(chapter4_table2, size = 8, part = "body")
chapter4_table2 <- fontsize(chapter4_table2, size = 8, part = "header")
chapter4_table2 <- colformat_num(chapter4_table2,digits = 1)
chapter4_table2 <- set_table_properties(chapter4_table2, layout = "autofit")
chapter4_table2 <- set_caption(chapter4_table2, "Table 2: Summary of the simulations run for the study. Simulations were run for each site over the course of 2 years and then pooled based on season for each year. The details pertaining are provided in the below. The start of each year was the beginning of Austral summer.") 
chapter4_table2
```

@@

To compare virtual kelp trajectories between sites the mean trajectories were calculated, and density distribution maps were produced. Density distribution maps are an effective method for assessing pathways and ocean connectivity [@vanSebille2018]. The distance particles traveled was also measured and used to produce boxplots. Distance traveled from the release site can help reveal topographic steering and momentum energy transfer from the wind field to kelp particles [@vanSebille2018]. If the wind direction is in the same direction as the current the virtual particles will cover larger distances due to an increase in the velocity because of momentum energy transfer [@putman2020; @putman2018; @vanSebille2018]. To aid in seasonal comparison of trajectories, proportions of each kelp phase at the end of each simulation run was calculated and compared. The different phases were surface, subsurface, beached, sunk and out of bounds.

# Results

```{r data, message=FALSE, warning=FALSE, include=FALSE}

total_summer <- read.csv(file = "OceanParcels/chapter_4_data/df_total_summer.csv") 
total_summer2 <- read.csv(file = "OceanParcels/chapter_4_data/df_total_summer2.csv")
total_autumn <- read.csv(file = "OceanParcels/chapter_4_data/df_total_autumn.csv") 
total_autumn2 <- read.csv(file = "OceanParcels/chapter_4_data/df_total_autumn2.csv")
total_winter <- read.csv(file = "OceanParcels/chapter_4_data/df_total_winter.csv") 
total_winter2 <- read.csv(file = "OceanParcels/chapter_4_data/df_total_winter2.csv")
total_spring <- read.csv(file = "OceanParcels/chapter_4_data/df_total_spring.csv") 
total_spring2 <- read.csv(file = "OceanParcels/chapter_4_data/df_total_spring2.csv")

# calculate and add overall velocity vector of ocean current sampled at virtual kelp location
total_summer$velocity <- with(total_summer, sqrt(VelU^2 + VelV^2))
total_summer2$velocity <- with(total_summer2, sqrt(VelU^2 + VelV^2))
total_autumn$velocity <- with(total_autumn, sqrt(VelU^2 + VelV^2))
total_autumn2$velocity <- with(total_autumn2, sqrt(VelU^2 + VelV^2))
total_winter$velocity <- with(total_winter, sqrt(VelU^2 + VelV^2))
total_winter2$velocity <- with(total_winter2, sqrt(VelU^2 + VelV^2))
total_spring$velocity <- with(total_spring, sqrt(VelU^2 + VelV^2))
total_spring2$velocity <- with(total_spring2, sqrt(VelU^2 + VelV^2))
# calculate and add overall kelp velocity vector sampled at virtual kelp location
total_summer$kelpvelocity <- with(total_summer, sqrt(KelpVelU^2 + KelpVelV^2))
total_summer2$kelpvelocity <- with(total_summer2, sqrt(KelpVelU^2 + KelpVelV^2))
total_autumn$kelpvelocity <- with(total_autumn, sqrt(KelpVelU^2 + KelpVelV^2))
total_autumn2$kelpvelocity <- with(total_autumn2, sqrt(KelpVelU^2 + KelpVelV^2))
total_winter$kelpvelocity <- with(total_winter, sqrt(KelpVelU^2 + KelpVelV^2))
total_winter2$kelpvelocity <- with(total_winter2, sqrt(KelpVelU^2 + KelpVelV^2))
total_spring$kelpvelocity <- with(total_spring, sqrt(KelpVelU^2 + KelpVelV^2))
total_spring2$kelpvelocity <- with(total_spring2, sqrt(KelpVelU^2 + KelpVelV^2))

total_summer_mean <- total_summer %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))
total_summer2_mean <- total_summer2 %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))

total_autumn_mean <- total_autumn %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))
total_autumn2_mean <- total_autumn2 %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))

total_winter_mean <- total_winter %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))
total_winter2_mean <- total_winter2 %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))

total_spring_mean <- total_spring %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))
total_spring2_mean <- total_spring2 %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))

#total_summer_mean <- read.csv(file = "OceanParcels/chapter_4_data/df_total_summer_mean.csv") 
#total_summer2_mean <- read.csv(file = "OceanParcels/chapter_4_data/df_total_summer2_mean.csv")
#total_autumn_mean <- read.csv(file = "OceanParcels/chapter_4_data/df_total_autumn_mean.csv") 
#total_autumn2_mean <- read.csv(file = "OceanParcels/chapter_4_data/df_total_autumn2_mean.csv")
#total_winter_mean <- read.csv(file = "OceanParcels/chapter_4_data/df_total_winter_mean.csv") 
#total_winter2_mean <- read.csv(file = "OceanParcels/chapter_4_data/df_total_winter2_mean.csv")
#total_spring_mean <- read.csv(file = "OceanParcels/chapter_4_data/df_total_spring_mean.csv") 
#total_spring2_mean <- read.csv(file = "OceanParcels/chapter_4_data/df_total_spring2_mean.csv")

total_summer_surface <- total_summer %>% 
filter(Phase == 1)
total_summer_subsurface <- total_summer %>% 
filter(Phase == 2)
total_summer_sink <- total_summer %>% 
filter(Phase == 3)
total_summer_beached <- total_summer %>% 
filter(Phase == 4)
total_summer_boundry <- total_summer %>% 
filter(Phase == 5)
total_autumn_surface <- total_autumn %>% 
filter(Phase == 1)
total_autumn_subsurface <- total_autumn %>% 
filter(Phase == 2)
total_autumn_sink <- total_autumn %>% 
filter(Phase == 3)
total_autumn_beached <- total_autumn %>% 
filter(Phase == 4)
total_autumn_boundry <- total_autumn %>% 
filter(Phase == 5)
total_winter_surface <- total_winter %>% 
filter(Phase == 1)
total_winter_subsurface <- total_winter %>% 
filter(Phase == 2)
total_winter_sink <- total_winter %>% 
filter(Phase == 3)
total_winter_beached <- total_winter %>% 
filter(Phase == 4)
total_winter_boundry <- total_winter %>% 
filter(Phase == 5)
total_spring_surface <- total_spring %>% 
filter(Phase == 1)
total_spring_subsurface <- total_spring %>% 
filter(Phase == 2)
total_spring_sink <- total_spring %>% 
filter(Phase == 3)
total_spring_beached <- total_spring %>% 
filter(Phase == 4)
total_spring_boundry <- total_spring %>% 
filter(Phase == 5)

total_summer2_surface <- total_summer2 %>% 
filter(Phase == 1)
total_summer2_subsurface <- total_summer2 %>% 
filter(Phase == 2)
total_summer2_sink <- total_summer2 %>% 
filter(Phase == 3)
total_summer2_beached <- total_summer2 %>% 
filter(Phase == 4)
total_summer2_boundry <- total_summer2 %>% 
filter(Phase == 5)
total_autumn2_surface <- total_autumn2 %>% 
filter(Phase == 1)
total_autumn2_subsurface <- total_autumn2 %>% 
filter(Phase == 2)
total_autumn2_sink <- total_autumn2 %>% 
filter(Phase == 3)
total_autumn2_beached <- total_autumn2 %>% 
filter(Phase == 4)
total_autumn2_boundry <- total_autumn2 %>% 
filter(Phase == 5)
total_winter2_surface <- total_winter2 %>% 
filter(Phase == 1)
total_winter2_subsurface <- total_winter2 %>% 
filter(Phase == 2)
total_winter2_sink <- total_winter2 %>% 
filter(Phase == 3)
total_winter2_beached <- total_winter2 %>% 
filter(Phase == 4)
total_winter2_boundry <- total_winter2 %>% 
filter(Phase == 5)
total_spring2_surface <- total_spring2 %>% 
filter(Phase == 1)
total_spring2_subsurface <- total_spring2 %>% 
filter(Phase == 2)
total_spring2_sink <- total_spring2 %>% 
filter(Phase == 3)
total_spring2_beached <- total_spring2 %>% 
filter(Phase == 4)
total_spring2_boundry <- total_spring2 %>% 
filter(Phase == 5)

total_summer_surface_mean <- total_summer_surface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))
total_summer_subsurface_mean <- total_summer_subsurface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))

total_summer2_surface_mean <- total_summer2_surface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))
total_summer2_subsurface_mean <- total_summer2_subsurface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))

total_autumn_surface_mean <- total_autumn_surface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))
total_autumn_subsurface_mean <- total_autumn_subsurface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))

total_autumn2_surface_mean <- total_autumn2_surface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))
total_autumn2_subsurface_mean <- total_autumn2_subsurface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))

total_winter_surface_mean <- total_winter_surface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))
total_winter_subsurface_mean <- total_winter_subsurface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))

total_winter2_surface_mean <- total_winter2_surface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))
total_winter2_subsurface_mean <- total_winter2_subsurface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))

total_spring_surface_mean <- total_spring_surface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))
total_spring_subsurface_mean <- total_spring_subsurface %>% # calculate mean trajectory
  group_by(t, site) %>%
  summarise(across(lon:kelpvelocity, mean, na.rm= TRUE))
